---
title: '#tidytuesday 19-01-22'
author: Elliot Palmer
date: '2019-01-21'
slug: tidytuesday-19-01-22
categories:
  - tidytuesday
tags: []
description: ''
featured: ''
featuredalt: ''
featuredpath: ''
linktitle: ''
type: post
---

# Introduction

[https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-01-22](Link to Github Repo)

From Github Readme:
>Data Info
>
>Data comes from [The Vera Institute GitHub](https://github.com/vera-institute/incarceration_trends). The raw dataset was >taken from their GitHub - it is in a wide format and if you are keen on really flexing your data munging skills it is a >worthy adversary! The truly raw data is seen >[here](https://raw.githubusercontent.com/vera-institute/incarceration_trends/master/incarceration_trends.csv). My full code >to reproduce the summary level datasets seen below can be found [here](example_code.R), you can adapt this minorly to get >more data from the original wide dataset.

## Loading Libraries

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(maps)
library(viridis)
```

## Read in Data

```{r get-data, message=FALSE, warning=FALSE, cache=TRUE}
# Setting the root in one place so that I don't have to necessarily repeat the path
repo_root <- "https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-01-22"

# This might take ~30s due to the volume of data
raw_data <- read_csv(glue::glue("{repo_root}/incarceration_trends.csv"))
pretrial_population <- read_csv(glue::glue("{repo_root}/pretrial_population.csv"))
pretrial_summary <- read_csv(glue::glue("{repo_root}/pretrial_summary.csv"))
prison_population <- read_csv(glue::glue("{repo_root}/prison_population.csv"))
prison_summary <- read_csv(glue::glue("{repo_root}/prison_summary.csv"))
```

## Supplementary Data

```{r}
state_name_to_abbrev <- data.frame(
  abbrev = state.abb,
  name = tolower(state.name)
)

county_xref <- raw_data %>% 
  select(fips, state, county_name) %>% 
  distinct() %>% 
  inner_join(state_name_to_abbrev, by = c("state" = "abbrev")) %>% 
  inner_join(maps::county.fips, by = "fips") %>% 
  tidyr::separate(polyname, c("state","county"), ",") %>% 
  select(-name)

# Mapping Data
states <- map_data("state") %>% 
  inner_join(state_name_to_abbrev, by = c("region" = "name"))

counties <- map_data("county") %>% 
  inner_join(state_name_to_abbrev, by = c("region" = "name")) %>% 
  left_join(county_xref, by = c("region" = "state", "subregion" = "county"))
  
```


## Data Munging

```{r data-munging}
prison_population <-
  prison_population %>% 
  mutate(pop_category_group = case_when(pop_category %in% c("Asian", "Black", "Latino", "Native American", "White", "Other") ~ "Ethnicity",
                                        pop_category %in% c("Female", "Male") ~ "Gender",
                                        pop_category == "Total" ~ "Total",
                                        TRUE ~ "Other")
         )

prison_pop_clean <- prison_population %>% 
  filter(between(year, 1990, 2015))
```


```{r location-data}
# State Data ----


state_prison_pop <- prison_pop_clean %>% 
  group_by(year, state) %>% 
  filter(pop_category_group == "Total") %>% 
  summarise(population = sum(population, na.rm = TRUE),
            prison_population = sum(prison_population, na.rm = TRUE)) %>% 
  mutate(prison_pop_per1k = prison_population / (population / 1000))

low_reporting_states <- state_prison_pop %>% 
  group_by(state) %>% 
  summarise(data_count = sum(prison_population > 0)) %>% 
  filter(data_count < 24) %>% 
  arrange(data_count)

state_prison_pop_median <- state_prison_pop %>% 
  filter(!(state %in% low_reporting_states$state)) %>% 
  group_by(state) %>% 
  summarise(median_rate = median(prison_pop_per1k)) 

# County Data ----
county_prison_pop <- prison_pop_clean %>% 
  group_by(year, state, county_name) %>% 
  filter(pop_category_group == "Total") %>% 
  summarise(population = sum(population, na.rm = TRUE),
            prison_population = sum(prison_population, na.rm = TRUE)) %>% 
  mutate(prison_pop_per1k = prison_population / (population / 1000))

county_prison_pop_median <- county_prison_pop %>% 
  filter(!(state %in% low_reporting_states$state)) %>% 
  group_by(state, county_name) %>% 
  summarise(median_rate = median(prison_pop_per1k)) 
```


### Exploration

Let's start with the prison population.
```{r}
# Let's Explore the different variables

prison_population %>% group_by(pop_category_group) %>% count()
```

# Data Validation

* Total Population reporting switched from `Black, White, Other` -> `Asian, Black, White, Latino, Native American` in 1990
  * Based on the dramatic drop in `White` counts, many `Latino`, `Native American`, and `Asian people we're miscategorized prior to 1990
  
* Prison Populations were reported differently by Ethnicity.
  * Prison Populations are recorded between 1983 and 2015
  * It appears that `Latino` was recorded in prison populations starting in 1983
  
* Therefore, we should restrict analysis from 1990 to 2015

* Many States (17) don't have prison population data for all years in the analysis

### Reporting of Ethnicity Changed in 1990
```{r}

prison_population %>% 
  #filter(complete.cases(.)) %>% 
  group_by(pop_category, year) %>% 
  filter(pop_category_group == "Ethnicity") %>% 
  summarise(population = sum(population, na.rm = TRUE),
            prison_population = sum(prison_population, na.rm = TRUE)) %>% 
  arrange(year) %>% 
  ggplot(aes(x = year, y = population, color = pop_category)) + 
  geom_line() +
  labs(title = "Population by Ethnicity",
       subtitle = "Reporting Changed in 1990",
       x = "Year",
       y = "Total Population") +
  scale_y_continuous(label = scales::comma) +
  theme_light()
  
```

```{r}
prison_population %>% 
  #filter(complete.cases(.)) %>% 
  group_by(pop_category, year) %>% 
  filter(pop_category_group == "Ethnicity") %>% 
  summarise(population = sum(population, na.rm = TRUE),
            prison_population = sum(prison_population, na.rm = TRUE)) %>% 
  arrange(year) %>% 
  ggplot(aes(x = year, y = prison_population, color = pop_category)) + 
  geom_line() +
  labs(title = "Prison Population by Ethnicity",
       subtitle = "Prison Pop Reporting Covers 1983 to 2015",
       x = "Year",
       y = "Prison Population") +
  scale_y_continuous(label = scales::comma) +
  theme_light()
```

### Some States have Spotty Reporting

```{r}
low_reporting_states
```


### Incarceration Rate by State
```{r}


state_prison_pop_median %>% 
  top_n(10, median_rate) %>% 
  ggplot(aes(x = forcats::fct_reorder(state, median_rate), y = median_rate)) +
  geom_point(size = 3) + 
  geom_linerange(aes(ymin = 0, ymax = median_rate), size = 2) +
  geom_label(aes(label = round(median_rate, 1)), nudge_y = .5) +
  labs(title = "Top 10 States by Median Incarceration Rate",
       subtitle = "1990 to 2015 | 25 Years | 17 of 50 States Excluded for Missing Data",
       x = "State", y = "Median Incarceration Rate (1990 - 2015)") +
  coord_flip() 

states %>% 
  filter(!(abbrev %in% low_reporting_states$state)) %>% 
  left_join(state_prison_pop_median, by = c("abbrev" = "state")) %>% 
  ggplot() + 
  geom_polygon(aes(x = long, y = lat, fill = median_rate , group = group), color = "black") + 
  scale_fill_continuous(na.value = "#FFFFFF",low = "white", high = "darkblue") +
  coord_fixed(1.3) +
  guides(fill=FALSE) +
  labs(title = "Map of Median Incarceration Rates",
       subtitle = "Missing States have no data | Gradiant from Low (Lightest) to High (Darkest)") +
  theme_void()# do this to leave off the color legend
```

```{r}
filter_county_prison_pop_median <- 
  county_prison_pop_median %>% 
  ungroup() %>% 
  mutate(median_rate_ntile = ntile(median_rate, 1000)) %>% 
  filter(between(median_rate_ntile, 2,998)) #%>% 

# All Counties
counties %>% 
  filter(!(abbrev %in% low_reporting_states$state)) %>% 
  left_join(filter_county_prison_pop_median, by = c("abbrev" = "state", "county_name")) %>% 
  ggplot() + 
  geom_polygon(aes(x = long, y = lat, group = group, fill = median_rate), color = "lightgray", size = .2)  +
  geom_polygon(data = states, aes(x = long, y = lat, group = group), color = "black", alpha = 0, size = .1) + 
  scale_fill_continuous(na.value = "#FFFFFF",low = "white", high = "darkblue") +
  coord_fixed(1.3) +
  guides(fill=FALSE) +
  labs(title = "Map of Median Incarceration Rates by County",
       subtitle = "Missing Counties / States have no data | Gradiant from Low (Lightest) to High (Darkest)") +
  coord_map() +
  theme_void()
```

```{r}
# Top Counties
counties %>% 
  filter(!(abbrev %in% low_reporting_states$state)) %>% 
  left_join(top_n(filter_county_prison_pop_median,500), by = c("abbrev" = "state", "county_name")) %>% 
  ggplot() + 
  geom_polygon(aes(x = long, y = lat, group = group, fill = median_rate), color = "lightgray", size = .2)  +
  geom_polygon(data = states, aes(x = long, y = lat, group = group), color = "black", alpha = 0, size = .1) + 
  scale_fill_continuous(na.value = "#FFFFFF",low = "white", high = "darkblue") +
  coord_fixed(1.3) +
  guides(fill=FALSE) +
  labs(title = "Map of Median Incarceration Rates by County: Top 500",
       subtitle = "Missing Counties / States have no data | Gradiant from Low (Lightest) to High (Darkest)") +
  coord_map() +
  theme_void()
```

## Data Analysis

## Conclusions